An R Markdown document converted from "01_Clustering.ipynb

### 2024-02-20 scripts/00_DataPrep.Rmd
Organise in folders

---
title: An R Markdown document converted from "11_DataEthnic.ipynb"
output: html_document
---

## Data from other ethnicities

@@

```{r}
tab_vars <- tribble(
    ~colid, ~colnam, ~coltyp,
    ## Basic
    "f.eid", "eid", "n",
    "f.53.0.0", "date0", "D",
    "f.31.0.0", "sex", "n",
    "f.21003.0.0", "age", "n",
    ## Anthropometric
    "f.21001.0.0", "bmi", "n",
    "f.48.0.0", "waist_circumference", "n",
    "f.49.0.0", "hip_circumference", "n",
    "f.50.0.0", "height", "n",
    ## Blood pressure
    "f.4080.0.0", "sbp_01", "n",
    "f.4080.0.1", "sbp_02", "n",
    "f.4079.0.0", "dbp_01", "n",
    "f.4079.0.1", "dbp_02", "n",
    ## Liver function
    "f.30620.0.0", "alt", "n",
    ## Kidney function
    "f.30700.0.0", "creatinine", "n",
    ## Inflammatory markers
    "f.30710.0.0", "crp", "n",
    ## Lipid fractions
    "f.30760.0.0", "hdl", "n",
    "f.30780.0.0", "ldl", "n",
    "f.30870.0.0", "tg", "n",
    ## Glycemia
    "f.30740.0.0", "glucose", "n",
    ## Smoking
    "f.1239.0.0", "current_smoker", "n",
    ## Caucasian ancestry
    "f.22006.0.0", "caucasian", "n",
    ## Self-reported ethnicity
    "f.21000.0.0", "srethnicity", "n"
)
```

@@

```{r}
nrow(crude_dat)
```

```{r}
recoded_ethn <- crude_dat %>%
    ## Only ancestries other than Caucasian
    filter(is.na(caucasian)) %>%
    select(-caucasian)
nrow(recoded_ethn)
```

```{r}
recoded_ethn$srethnicity %>% table
```

```{r}
ethnicitytab <- tribble(
    ~Ethnicity, ~srethnicity,
    "SouthAsian", c(3001, 3002, 3003),
    "African", c(4, 4001, 4002, 4003),
    "Chinese", c(5)
) %>%
    unnest(srethnicity)
ethnicitytab
```

```{r}
recoded_ethn <- recoded_ethn %>%
    inner_join(ethnicitytab, by = "srethnicity")
nrow(recoded_ethn)
```

```{r}
recoded_ethn <- recoded_ethn %>%
    transmute(eid, age, Ethnicity,
              sex = ifelse(sex == 0, "Female", "Male"),
              ## Anthropometric
              bmi, 
              whr = waist_circumference / hip_circumference,
              ## Blood pressure
              sbp = rowMeans(cbind(sbp_01, sbp_02), na.rm = TRUE),
              dbp = rowMeans(cbind(dbp_01, dbp_02), na.rm = TRUE),
              ## Liver function
              alt,
              ## Kidney function
              scr = creatinine,
              ## Inflammatory markers
              crp,
              ## Lipid fractions
              hdl, tg, ldl,
              ## Glycemia - At least fasting time of 4 hours
              fg = ifelse(fasting_time >= 4, glucose, NaN),
              ## Current smoking status
              smoking = case_when(current_smoker %in% c(1,2) ~ 1,
                                  current_smoker == 0 ~ 0,
                                  TRUE ~ NaN)) %>%
    ## Only complete cases
    drop_na
head(recoded_ethn)
nrow(recoded_ethn)
```

```{r}
table(recoded_ethn$Ethnicity)
```

```{r}
recoded_ethn <- recoded_ethn %>%
    group_by(Ethnicity) %>%
    filter(n() >= 1000) %>%
    ungroup
table(recoded_ethn$Ethnicity)
```

## Covariate data

```{r}
covar_dat <- read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
head(covar_dat)
```

@@

```{r}
survmacedat <- mace_firstoccur %>%
    transmute(
        eid,
        outcome_value = 1, 
        outcome_date = event_date
    ) %>%
    right_join(datedat, by = "eid") %>%
    filter(eid %in% recoded_ethn$eid) %>%
    mutate(
        outcome_value = replace_na(outcome_value, 0),
        outcome_date = coalesce(outcome_date, censordate),
        event_after_censor = outcome_value == 1 & outcome_date > lastdate,
        outcome_value = if_else(event_after_censor, 0, outcome_value),
        outcome_date = if_else(event_after_censor, lastdate, outcome_date),
        outcome_timeyrs = as.numeric(outcome_date - date0) / 365.25
    ) %>%
    filter(outcome_date > date0) %>%
    select(-c(date0, outcome_date, lastdate, event_after_censor))
head(survmacedat)
nrow(survmacedat)
```

@@

@@

```{r}
primcareinds <- primcareinds %>%
    filter(eid %in% recoded_ethn$eid)
nrow(primcareinds)
```

@@

```{r}
primcareinds <- primcareinds %>%
    inner_join(covar_dat[,c("eid", "T1D", "T2D", "Insulin", "AntiDM")], by = "eid")
head(primcareinds)
nrow(primcareinds)
```

@@

@@

```{r}
write_tsv(recoded_ethn, "../data/recoded_ethn.tsv")
```

@@

```{r}
write_tsv(survmacedat, "../data/survmacedat_ethn.tsv")
```

@@

```{r}
write_tsv(survdmdat, "../data/survdmdat_ethn.tsv")
```

>>>>

---
title: An R Markdown document converted from "00_DataPrep.ipynb"
output: html_document
---

# Data preparation

## Libraries

@@

```{r}
tab_vars <- tribble(
    ~colid, ~colnam, ~coltyp,
    ## Basic
    "f.eid", "eid", "n",
    "f.53.0.0", "date0", "D",
    "f.31.0.0", "sex", "n",
    "f.21003.0.0", "age", "n",
    ## Anthropometric
    "f.21001.0.0", "bmi", "n",
    "f.48.0.0", "waist_circumference", "n",
    "f.49.0.0", "hip_circumference", "n",
    "f.50.0.0", "height", "n",
    ## Blood pressure
    "f.4080.0.0", "sbp_01", "n",
    "f.4080.0.1", "sbp_02", "n",
    "f.4079.0.0", "dbp_01", "n",
    "f.4079.0.1", "dbp_02", "n",
    ## Liver function
    "f.30620.0.0", "alt", "n",
    ## Kidney function
    "f.30700.0.0", "creatinine", "n",
    ## Inflammatory markers
    "f.30710.0.0", "crp", "n",
    ## Lipid fractions
    "f.30760.0.0", "hdl", "n",
    "f.30780.0.0", "ldl", "n",
    "f.30870.0.0", "tg", "n",
    ## Glycemia
    "f.30740.0.0", "glucose", "n",
    ## Smoking
    "f.1239.0.0", "current_smoker", "n",
    ## Caucasian ancestry
    "f.22006.0.0", "caucasian", "n"
)
```

@@

Recoding data:

```{r}
recoded_dat <- crude_dat %>%
    ## Only European descent individuals
    filter(caucasian == 1) %>%
    select(-caucasian) %>%
    transmute(eid, age, 
              sex = ifelse(sex == 0, "Female", "Male"),
              ## Anthropometric
              bmi, 
              whr = waist_circumference / hip_circumference,
              ## Blood pressure
              sbp = rowMeans(cbind(sbp_01, sbp_02), na.rm = TRUE),
              dbp = rowMeans(cbind(dbp_01, dbp_02), na.rm = TRUE),
              ## Liver function
              alt,
              ## Kidney function
              scr = creatinine,
              ## Inflammatory markers
              crp,
              ## Lipid fractions
              hdl, tg, ldl,
              ## Glycemia - At least fasting time of 4 hours
              fg = ifelse(fasting_time >= 4, glucose, NaN),
              ## Current smoking status
              smoking = case_when(current_smoker %in% c(1,2) ~ 1,
                                  current_smoker == 0 ~ 0,
                                  TRUE ~ NaN)) %>%
    ## Only complete cases
    drop_na
head(recoded_dat)
nrow(recoded_dat)
```

## Pre-existing conditions

```{r}
tab_vars <- tibble(
    colid = c("f.eid",  "f.31.0.0"),
    colnam = c("eid", "sex"),
) %>%
    bind_rows(
        data.frame(colid = paste0("f.20002.0.", 0:33),
                   colnam = paste("noncancercode", 1:34, sep = "_")),
        data.frame(colid = paste0("f.20009.0.", 0:33),
                   colnam = paste("noncancerage", 1:34, sep = "_"))
    )
tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
    pipe %>%
    read_table(col_names = "colid", col_types = "c") %>%
    ## Position of columns in table
    mutate(colpos = row_number()) %>%
    ## Extracting position and name of variables of interest
    inner_join(tab_vars, by = "colid")
head(tab_vars)
```

```{r}
dx_dat <- read_tsv(
    file = pipe(column_cmd),
    ## Replacing column header with variable names
    skip = 1, col_names = tab_vars$colnam,
    ## All columns are numeric
    col_types = cols(.default = "n")
)
head(dx_dat)
```

```{r}
nrow(dx_dat)
```

Long format

```{r}
dx_dat2 <- dx_dat %>%
    pivot_longer(
        -c(eid, sex),
        names_sep = "_",
        names_to = c(".value", NA),
        values_drop_na = TRUE
    )
head(dx_dat2)
```

Joining with coding:

```{r}
dx_coding <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/codings/coding6.tsv",
                      show_col_types = FALSE)
head(dx_coding)
```

```{r}
dx_dat2 <- dx_dat2 %>%
    inner_join(dx_coding[, c("coding", "meaning")],
               by = c("noncancercode" = "coding"))
head(dx_dat2)
```

### Selection of pre-existing conditions

```{r}
selected_dx <- dx_dat2 %>%
    group_by(eid) %>%
    summarise(
        HT = any(meaning %in% c("hypertension", 
                                "essential hypertension")),
        CHD = any(meaning %in% c("heart/cardiac problem",
                                 "angina", "heart attack/myocardial infarction")),
        Stroke = any(meaning %in% c("stroke", "ischaemic stroke",
                                    "transient ischaemic attack (tia)",
                                    "subarachnoid haemorrhage")),
        PAD = any(meaning %in% c("arterial embolism", "peripheral vascular disease",
                                 "leg claudication/ intermittent claudication")),
        CKD = any(meaning %in% c("renal/kidney failure",
                                 "renal failure requiring dialysis",
                                 "renal failure not requiring dialysis")),
        LiverFailure = any(meaning %in% c("liver failure/cirrhosis",
                                          "primary biliary cirrhosis",
                                          "alcoholic liver disease / alcoholic cirrhosis")),
        RA = any(meaning == "rheumatoid arthritis"),
        T2D = any(meaning %in% c("diabetes", "type 2 diabetes", "diabetic eye disease", 
                                 "diabetic neuropathy/ulcers")),
        T1D = any(meaning %in% c("type 1 diabetes")),
        T2Dage = noncancerage[
            meaning %in% c("diabetes", "type 2 diabetes", "diabetic eye disease", 
                           "diabetic neuropathy/ulcers")
        ][1]
    ) %>%
    mutate(across(-c(eid, T2Dage), ~.x * 1)) %>%
    right_join(dx_dat[,"eid"], by = "eid") %>%
    mutate(
        across(-eid, ~replace_na(.x, 0)),
        T2Dage = ifelse(T2Dage < 0, NaN, T2Dage),
        T1D = ifelse(T2D == 1 & T2Dage < 35, 1, T1D),
        T2D = ifelse(T1D == 1, 0, T2D),
        T2Dage = ifelse(T2D == 1 & T1D == 0, T2Dage, 0)
    ) %>%
    drop_na
head(selected_dx)
nrow(selected_dx)
```

## Medication data

```{r}
tab_vars <- tibble(
    colid = "f.eid",
    colnam = "eid",
) %>%
    bind_rows(
        data.frame(colid = paste0("f.20003.0.", 0:47),
                   colnam = paste("medcode", 1:48, sep = "_"))
    )
tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
    pipe %>%
    read_table(col_names = "colid", col_types = "c") %>%
    ## Position of columns in table
    mutate(colpos = row_number()) %>%
    ## Extracting position and name of variables of interest
    inner_join(tab_vars, by = "colid")
head(tab_vars)
```

```{r}
meds_dat <- read_tsv(
    file = pipe(column_cmd),
    ## Replacing column header with variable names
    skip = 1, col_names = tab_vars$colnam,
    ## All columns are numeric
    col_types = cols(.default = "n")
)
head(meds_dat)
```

Long format

```{r}
meds_dat2 <- meds_dat %>%
    pivot_longer(
        -eid,
        names_sep = "_",
        names_to = c(".value", NA),
        values_drop_na = TRUE
    )
head(meds_dat2)
```

Joining with coding:

```{r}
meds_coding <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/codings/coding4.tsv",
                      show_col_types = FALSE)
head(meds_coding)
```

```{r}
meds_dat2 <- meds_dat2 %>%
    inner_join(meds_coding,
               by = c("medcode" = "coding"))
head(meds_dat2)
```

Map of UK Biobank medication to ATC codes (available for some medications only):

```{r}
ATC_matches <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/codings/ATC_matches.tsv", 
                        col_types = "ncc")
head(ATC_matches)
```

Selected codes to use:

- A10A - Insulin
- A10B - Diabetes drugs
- C01 - Cardiac therapy
- C02 - Antihypertensives
- C03 - Diuretics
- C07 - Beta blocking agents
- C08 - Calcium channel blockers
- C09 - RA system drugs
- C10 - Lipid lowering

```{r}
ATC_matches_sel <- ATC_matches %>%
    filter(grepl("^A10|^C01|^C02|^C03|^C07|^C08|^C09|^C10", ATC_codes)) %>%
    unique
head(ATC_matches_sel)
```

```{r}
selmeds <- meds_dat2 %>%
    inner_join(ATC_matches_sel, by = c("medcode" = "ukb_code", 
                                       "meaning" = "med_name"))
head(selmeds)
```

Ocurrence of selected medication:

```{r}
selected_meds <- selmeds %>%
    group_by(eid) %>%
    summarise(
        Insulin = any(ATC_codes %in% c("A10A")),
        AntiDM = any(ATC_codes %in% c("A10B")),
        AntiHT = any(grepl("^C01|^C02|^C03|^C07|^C08|^C09", ATC_codes)),
        LipidLower = any(grepl("^C10", ATC_codes))
    ) %>%
    mutate(across(-eid, ~.x * 1)) %>%
    right_join(meds_dat[,"eid"], by = "eid") %>%
    mutate(across(-eid, ~replace_na(.x, 0)))
head(selected_meds)
nrow(selected_meds)
```

## Joining pre-existing conditions and medication data

```{r}
covar_dat <- inner_join(selected_dx, selected_meds, by = "eid")
head(covar_dat)
nrow(covar_dat)
```

@@

```{r}
survmacedat <- mace_firstoccur %>%
    transmute(
        eid,
        outcome_value = 1, 
        outcome_date = event_date
    ) %>%
    right_join(datedat, by = "eid") %>%
    filter(eid %in% recoded_dat$eid) %>%
    mutate(
        outcome_value = replace_na(outcome_value, 0),
        outcome_date = coalesce(outcome_date, censordate),
        event_after_censor = outcome_value == 1 & outcome_date > lastdate,
        outcome_value = if_else(event_after_censor, 0, outcome_value),
        outcome_date = if_else(event_after_censor, lastdate, outcome_date),
        outcome_timeyrs = as.numeric(outcome_date - date0) / 365.25
    ) %>%
    filter(outcome_date > date0) %>%
    select(-c(date0, outcome_date, lastdate, event_after_censor))
head(survmacedat)
nrow(survmacedat)
```

@@

We will extract information contained in primary care data to assess who developed diabetes after recruitment. The first thing needed is the set of individuals for which primary care data is available - not everyone has data available:

```{r}
system(
    paste(
        "cut", "-f", 1, 
        "/ludc/Home/daniel_c/projects/DVA/Data/UKBiobank/ukb_57232/gp_data/gp_clinical.txt", "|",
        "tail", "-n", "+2", "|",
        "sort", "|",
        "uniq", ">", "../data/primcareinds.tsv"
    )
)
```

@@

```{r}
primcareinds <- primcareinds %>%
    filter(eid %in% recoded_dat$eid)
nrow(primcareinds)
```

@@

```{r}
primcareinds <- primcareinds %>%
    inner_join(selected_dx[,c("eid", "T1D", "T2D")], by = "eid") %>%
    inner_join(selected_meds[,c("eid", "Insulin", "AntiDM")])
head(primcareinds)
nrow(primcareinds)
```

@@

```{r}
primcarecodes_dm <- read_tsv("../data/PrimaryCareCodesDM.txt", show_col_types = FALSE)
head(primcarecodes_dm)
```

Saving only the codes to look for them in the primary care table:

```{r}
primcarecodes_dm %>%
    select(Code) %>%
    write_tsv("../data/PrimCareCodesDM.tsv", col_names = FALSE)
```

Looking for codes in primary care data table:

```{r}
system(
    paste(
        "grep", "-Fwf", "../data/PrimCareCodesDM.tsv",
        "/ludc/Home/daniel_c/projects/DVA/Data/UKBiobank/ukb_57232/gp_data/gp_clinical.txt", ">",
        "../data/DM_events_PrimCare.tsv"
    )
)
```

@@

```{r}
write_tsv(recoded_dat, "../data/recoded_dat.tsv")
```

~~~~~~~~~~~~~~~~~~~~~~~

Pre-existing conditions and medication data:

```{r}
write_tsv(covar_dat,  "../data/covar_dat.tsv")
```

~~~~~~~~~~~~~~~~~~~~~~~

@@

```{r}
write_tsv(survmacedat, "../data/survmacedat.tsv")
```

@@

```{r}
write_tsv(survdmdat, "../data/survdmdat.tsv")
```

### 2024-02-20 scripts/01_Clustering.Rmd
Organise in folders

```{r}
source("cross_sectional_FX.R")
```

@@

@@

## Saving discovery cohort used

@@

>>>>

```{r}
source("01_cross_sectional_FX.R")
```

@@

The `umap_model` object contains the necessary elements to map new individuals to an existing embedding, we can extract this object with the following function:

```{r}
umap_models <- get_umap_mod(cluster_results)
```

This object may be useful for future analyses. However, this object may contain individual data. Additionally, for the current analyses the only object we require from `umap_model` from other cohorts is the embedding for visual assessment. Therefore, we have the following function so that in `umap_model` only the embedding is retained:

```{r}
cluster_results <- clean_cluster_results(cluster_results)
```

@@

## Saving other objects generated

Other objects generated that we would ask to save for future analyses, but we won't ask you to upload:

- The inputa data, not residualised:

@@

- The input data, residualised:

```{r}
save(residtab, file = "../data/ukb/residtab.RData")
```

- The UMAP models:

```{r}
iwalk(
    umap_models,
    ~save_uwot(
        model = .x, 
        file = paste0("/ludc/Home/daniel_c/projects/DVA/CrossSectionalBMI/data/ukb/umapmod_", .y)
    )
)
```
